Navigating Through Crisis: The Influence of COVID-19 on Canada’s Mental Health Problem

### 2024-04-18 paper/paper.qmd
Updated

The global COVID-19 pandemic has had a profound impact on public health, particularly in the realm of mental health. In Canada, the challenges brought on by the pandemic have intensified existing mental health issues and revealed significant vulnerabilities within the healthcare system (Government of Canada, 2021) *cite this*. This paper explores the effects of the COVID-19 pandemic on mental health trends in Canada, with a particular focus on the shifts in prevalence and the disparities in healthcare access observed across different sociodemographic groups during this period.

Research on the long-term mental health trends during the COVID-19 pandemic is sparse, particularly studies that provide a comprehensive analysis across various sociodemographic groups. Most existing studies have concentrated on immediate or short-term impacts within narrowly defined populations. This study aims to bridge this gap by conducting a detailed examination of how factors such as age, gender, and geographic location have influenced mental health outcomes across the broader Canadian population during the pandemic (CAMH, 2020) *cite this*.

The methodology of this paper includes a comparative analysis of data from national health surveys, focusing on the period from 2019 to the latest available data in 2021. This approach allows us to observe both pre-pandemic and during-pandemic mental health indicators, identifying significant trends and changes. Preliminary observations suggest that there are significant disparities in the impact of the pandemic on mental health, which vary widely among different demographic groups.

This research is critical not only for academic purposes but also for policymakers and healthcare providers planning interventions for ongoing and future public health crises. The structure of the paper is as follows: Section 1 introduces the background and literature relevant to the study; Section 2 outlines the methodology and data sources; Section 3 presents the analysis of the findings; Section 4 discusses the implications of these findings for mental health policy and practice in Canada.

@@

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

>>>>

The COVID-19 pandemic has spotlighted mental health as a critical component of public health, magnifying existing challenges and uncovering gaps within healthcare systems globally. In Canada, the impact of the pandemic on mental health has been profound, exacerbating disparities and highlighting the urgency of effective mental health care. The Centre for Addiction and Mental Health (CAMH) has identified a significant increase in mental health problems among Canadians, noting that the pandemic has led to unprecedented levels of anxiety and depression (CAMH, 2020). Additionally, the Public Health Agency of Canada reports on the elevated symptoms of anxiety and depression during the pandemic, further underscoring the need for comprehensive data analysis on this issue (Public Health Agency of Canada, 2021).

This paper aims to provide a detailed analysis of the changing situation of mental health in Canada by using four datasets spanning from 2002 to 2023. The focus is on perceived mental health statuses across various demographic segments, including age, gender, and province. The primary objective is to understand how perceptions of mental health have evolved over the past two decades and how these perceptions have been impacted by the COVID-19 pandemic.

The study addresses the gap in longitudinal mental health research, particularly in the analysis of sociodemographic influences during a major global crisis. Through the comparison of mental health indicators before and during the pandemic, this research provides insights into the shifts in mental health perceptions and the effectiveness of the responses to these changes. Initial findings suggest notable disparities in mental health impacts across different demographic groups, indicating the need for targeted interventions.

The significance of this research lies in its contribution to the evidence base informing mental health policy and practice. By identifying the demographic factors associated with varying mental health outcomes, the study aids policymakers and healthcare providers in developing tailored strategies to address the unique challenges faced by diverse groups within Canada. 
*TBD*

@@

This study utilizes four carefully curated datasets from Statistics Canada, each providing valuable insights into the mental health perceptions of Canadians. These datasets are pivotal for understanding the evolution of mental health across different demographics and geographical regions from 2002 to 2023.

## Overview of Datasets

The data for this research come from national surveys conducted by Statistics Canada, specifically designed to monitor various aspects of mental health across the country. Each dataset targets different demographic variables, such as age, gender, and province, and covers distinct periods critical for analyzing long-term trends and the immediate effects of the COVID-19 pandemic.

### Mental Health Indicators Dataset
Data from 2002 and 2012 in the Mental Health Indicators Dataset were derived from national health databases, capturing key statistics on the prevalence of mental disorders and utilization of mental health services. The information is gathered through comprehensive health surveys and administrative records, providing robust measures of mental health status and service outcomes.

### Perceived Mental Health by Age Group
The Perceived Mental Health by Age Group dataset spans from 2015 to 2022 and utilizes a consistent methodological framework to assess self-reported mental health. The data collection process involves standardized questionnaires distributed to a representative sample of Canadian adults across various age groups. This approach ensures that the data reflect a broad spectrum of the population, excluding those in full-time residence, on Indigenous reserves, or non-permanent residents, thus focusing on a generalizable sample of the Canadian populace.

### Perceived Mental Health by Gender and Sociodemographic Characteristics
The data for 2021 to 2023 from this dataset is sourced from the Canadian Community Health Survey (CCHS), which meticulously gathers data on mental health perceptions influenced by gender and other sociodemographic factors such as income, education, and employment. This survey selects participants through a stratified sampling design aimed at capturing diverse sociodemographic landscapes across Canada, enhancing the dataset's relevance for analyzing intersectional impacts on mental health.

## Detailed Measurement Discussion

Statistics Canada employs rigorous methodologies to ensure data accuracy and reliability. For example, the survey instruments are tested extensively before deployment to ensure they accurately capture the intended data. Additionally, data collectors are trained to adhere to strict guidelines to maintain consistency and reduce potential biases during data collection.

The sample for each dataset is designed to be representative of the Canadian population, stratified by key demographic variables to ensure comprehensive coverage across different regions and groups. This stratification helps in understanding specific sub-population dynamics within the broader national context. Data is then weighted to adjust for potential non-response biases and to represent the population accurately.

Despite these rigorous methods, the reliability of self-reported data can sometimes be challenging to assess due to the subjective nature of mental health perceptions. Statistics Canada addresses this by implementing multiple stages of data verification and validation, including automated data checks and manual reviews by experts. Furthermore, unusual or outlier responses undergo additional scrutiny to confirm their validity.
